\chapter{Source code}
\label{ch:code}

The code written for this thesis is open-source. For each figure and table we
provide a Python script that reproduces the content. Each script is referenced
at the end of the caption like this: \scriptlink{fignoise.py}. In the PDF it
links is to a preview of the file in an online repository,
\url{https://github.com/Gattocrucco/sipmfilter}. To run these scripts, clone or
download the repository. The scripts are located in the directory
\nolinkurl{figthesis}. The file \nolinkurl{README.md} provides detailed
instructions on how to set up the working environment.

Some scripts require no input. Others require large data files which are not
included in the repository. Of the latter, some have a cache of the data they
need, others just can not be executed without the original data. For people
outside of the DarkSide collaboration it may not be possible to obtain the data
files; in any case, they can check what the code is doing exactly.

The \LaTeX{} code for this thesis is itself available at
\url{https://github.com/Gattocrucco/thesis}. Moreover, the
\nolinkurl{sipmfilter} repository contains slides on this work which we
presented at DarkSide meetings, although they shall be considered outdated
respect to this document. The Python code is covered by the MIT license, while
the thesis and the slides are covered by the CC-BY~4.0. These licenses grant
anyone the right to reuse this material, even without releasing themselves the
modifications as open source, provided they cite the original author and keep
the copyright and license notices. The licenses do not cover figures which we
copied from other sources nor the DarkSide data.
